Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation

### 2014-05-08 doc/ttsp2013/ttsp_2013.tex
rearranging the mcsa and spn sections in the ttsp paper

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
\\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}}
}

@@

We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
solver technique for the simplified $P_N$ ($SP_N$) discretization of
the Boltzmann neutron transport equation. Using the fully-formed
linear operator for the transport problem, we explore solutions to the
$SP_N$ equations with MCSA using a challenging light water reactor
fuel assembly criticality calculation as the driving problem. Due to
neutron scattering in the light water moderator, several difficulties
arise when applying MCSA to these problems that were not observed in
previous work where the method was applied to more opaque thermal
radiaton transport systems. Of most significance, it was discovered
that simply reducing the spectral radius of the Monte Carlo problem
below unity is not enough to achieve convergence as outlined in
previous work. In order to effectively solve the $SP_N$ equations for
the model problem, an algebraic preconditioning strategy is developed
for MCSA where we utilize incomplete factorizations as preconditioners
to achieve convergence. Using the preconditioners, MCSA performance is
compared with production implementations of the preconditioned GMRES
and BiCGSTAB methods. Studies indicate that the iterative performance
of the method is competitive with the preconditioned Krylov methods
when operator spectrum is properly conditioned.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization, and high
confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems are often variants on
the discrete ordinates ($S_N$) method \cite{evans_denovo:_2010}. For
fission reactor neutronics simulations, the $S_N$ method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. In the 1960's, Gelbard developed
an ad-hoc, multidimensional extension of the simple single dimension
planar $P_N$ equations that created a system of coupled,
diffusion-like equations known as the simplified $P_N$ ($SP_N$)
equations (cite gelbard). The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required and permit larger simulations to be performed on the
available computing platforms.

Looking towards future computing platforms, as machines begin to
operate at hundreds of petaflops peak performance and beyond, trends
toward reduced energy consumption will require incredibly high levels
of concurrency to achieve the desired computation rates. The end
result of these hardware changes is that the larger number of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events at the application level. With solutions to
deterministic transport problems based on Monte Carlo techniques, such
issues are potentially alleviated by statistical arguments. In the
case of soft failures, isolated floating point errors in the Monte
Carlo simulation are absorbed within tally statistics while a hard
failure is treated as a high variance event where some portion of the
Monte Carlo histories lost.

Considering these physics and hardware issues, the goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. In previous work \cite{evans_monte_2012}, we
applied the Monte Carlo synthetic acceleration (MCSA) method to the
thermal radiation diffusion equation. For several transient problems
we observed that MCSA performed competetively with preconditioned
Krylov methods. In a continuation of these studies, we apply the
method to light water reactor problems discretized by the $SP_N$
equations. This system has no time dependence, a large scattering
component, and assymetry in the linear operator, creating a set of
problems complementary to the transient, opaque, and symmetric systems
of our previous analyis. In addition, we apply a new fixed point
iteration in the MCSA sequence as well as a new estimator in order to
study their effects on convergence.

The paper is organized as follows. First, we briefly introduce the
model $SP_N$ problem and describe the application of the MCSA
algorithm. Next, we demonstrate that the basic Jacobi preconditioning
applied in our previous work is unsatisfactory to achieve convergence
for neutron transport in light water reactors and subsquently develop
a new preconditioning scheme. We then apply this preconditioning
scheme to a light water reactor fuel assembly and compare the
performance of the preconditioned MCSA method to preconditioned Krylov
methods.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}):
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$. We apply the following change of
variables to yield the set of multigroup pseudo-moments
$\mathbb{U}_n$:
\begin{subequations}
  \begin{gather}
    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
where now the application of the fission matrix to the moment vectors
will be expanded into a set of block matrices in identical fashion to
the scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in
\cite{evans_simplified_2013}. It should be noted here that with
respect to a general MCSA scheme, the introduction of fission in the
system will only affect the source vector in the linear system for the
eigenvalue schemes used in this work. The linear operator, and
therefore overall MCSA performance will be dictated by streaming and
scattering as defined on the left-hand side.

\subsection{Monte Carlo Synthetic Acceleration}
\label{subsec:mcsa}
In this section we briefly review the Monte Carlo method for linear
systems and Monte Carlo Synthetic Acceleration. Given a linear
problem:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
\mathbb{R}^N$. To form the Monte Carlo method we define the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
following adjoint inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
We can acquire the adjoint solution by forming the Neumann series by
writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence by forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

In addition to the collision estimator, an additional estimator is
available due to the work of Okten \cite{okten_solving_2005} that uses
the method of expected values as a means to improve the Monte Carlo
estimate. As outlined by Spanier and Gelbard
\cite{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator}\footnote{The variance for this estimator is discussed in
  Appendix~\ref{chap:estimator_variance}.} for a random walk of $k$
events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potential states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to the
collision estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore an unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

At each eigenvalue iteration, we then solve the linear system with the
MCSA:
\begin{subequations}
  \begin{gather}
    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
    \mathbf{\Phi}_{l+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the $l$ subscript indicates the MCSA iteration number and $n$
again indicates the eigenvalue iteration. We refer the reader to
\cite{evans_monte_2012} for more a more complete definition of the
Monte Carlo method for linear systems and MCSA.

@@

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their similar diffusion-like form. A single energy
group problem was first computed with $SP_1$ discretization,
effectively giving the one-speed neutron diffusion system for the fuel
assembly resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.

>>>>

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton, Paul
  P.H. Wilson
  \\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}} }

@@

We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
solver technique for the simplified $P_N$ ($SP_N$) discretization of
the Boltzmann neutron transport equation. Using the fully-formed
linear operator for the transport problem, we explore solutions to the
$SP_N$ equations with MCSA using a challenging light water reactor
fuel assembly criticality calculation as the driving problem. Due to
neutron scattering in the light water moderator, several difficulties
arise when applying MCSA to these problems that were not observed in
previous work where the method was applied to more opaque thermal
radiaton transport systems. Of most significance, it was discovered
for these problems that simply reducing the spectral radius of the
Monte Carlo problem below unity is not enough to efficiently achieve
convergence as reported in previous work. In order to effectively
solve the $SP_N$ equations for the model problem, an algebraic
preconditioning strategy is developed for MCSA where we utilize
incomplete factorizations as preconditioners to achieve
convergence. Using the preconditioners, MCSA performance is compared
with production implementations of the preconditioned GMRES and
BiCGSTAB methods. Our studies indicate that the iterative performance
of MCSA when used to solve the $SP_N$ equations for light water
reactor problems is competitive with the preconditioned Krylov methods
when operator spectrum is properly conditioned.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization and burnup, and
high confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems where the solution over
the entire core of a reactor is desired are often variants on the
discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
method of characteristics \cite{askew_1972}. For fission reactor
neutronics simulations, the discrete ordinates method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011} which in many cases can be
intractable without sufficient computational hardware. On the other
side of the spectrum, low-fidelity diffusion solutions are often
leveraged as an alternative to compute a solution over the entire
domain a reactor core. In the 1960's, Gelbard developed an ad-hoc,
multidimensional extension of the simple single dimension planar $P_N$
equations that created a system of coupled, diffusion-like equations
known as the simplified $P_N$ ($SP_N$) equations \cite{gelbard_1960}
which incorporates additional angular information into a set of
diffusion-like equations. Further work in both aysmptotic analysis and
algorithmic development proceeded in the following decades to serve as
a theoretical basis for the method
\cite{morel_1996,brantley_simplified_2000}. Recent work has expanded
the discretization to a fully algebraic form ammenable to solution
with modern linear algebra techniques including Krylov solvers and
advanced precondtioners including multigrid methods and incomplete
factorizations \cite{hamilton_2014}. Using these new techniques, the
reduction in numerical complexity compared to current deterministic
solution methods using the $SP_N$ approximation could mean significant
savings in both compute time and memory required and permit larger
simulations to be performed on the available computing platforms.

Looking towards future computing platforms, as machines begin to
operate at hundreds of petaflops peak performance and beyond, trends
toward reduced energy consumption will require incredibly high levels
of concurrency to achieve the desired computation rates
\cite{kogge_using_2011}. The end result of these hardware changes is
that the larger number of low-powered processors will be prone to both
soft failures such as bit errors in floating point operations and hard
failures where the data owned by that processor cannot be recovered
\cite{u.s._department_of_energy_resilient_2012}. Because these
failures are predicted to be common, resilient solver technologies are
are a possible means of overcoming these events at the application
level. With solutions to deterministic transport problems based on
Monte Carlo techniques, such issues are potentially alleviated by
statistical arguments. In the case of soft failures, isolated floating
point errors in the Monte Carlo simulation are absorbed within tally
statistics while a hard failure is treated as a high variance event
where some portion of the Monte Carlo histories lost.

Considering these physics and hardware issues, the goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. In previous work \cite{evans_monte_2014}, we
applied the Monte Carlo synthetic acceleration (MCSA) method to the
thermal radiation diffusion equation. For several transient problems
we observed that MCSA performed competetively with preconditioned
Krylov methods. In a continuation of these studies, we apply the
method to light water reactor problems discretized by the $SP_N$
equations. This system has no time dependence, a large scattering
component, and assymetry in the linear operator, creating a set of
problems complementary to the transient, opaque, and symmetric systems
of our previous analyis. In addition, we apply a new fixed point
iteration in the MCSA sequence as well as a new estimator in order to
study their effects on convergence.

The paper is organized as follows. First, we briefly introduce the
$SP_N$ equations for eigenvalue problems along with the discretization
used for this work and the model problem. We then describe the
application of the MCSA algorithm. Next, we demonstrate that the basic
Jacobi preconditioning applied in our previous work is unsatisfactory
to achieve convergence for neutron transport in light water reactors
and subsquently develop a new preconditioning scheme. We then apply
this preconditioning scheme to a light water reactor fuel assembly and
compare the performance of the preconditioned MCSA method to
preconditioned Krylov methods followed by a summary and conclusion.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}):
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$. We apply the following change of
variables to yield the set of multigroup pseudo-moments
$\mathbb{U}_n$:
\begin{subequations}
  \begin{gather}
    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
where now the application of the fission matrix to the moment vectors
will be expanded into a set of block matrices in identical fashion to
the scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in \cite{hamilton_2014} such
that the flux and current are continuous at cell boundaries.. It
should be noted here that with respect to a general MCSA scheme, the
introduction of fission in the system will only affect the source
vector in the linear system for the eigenvalue schemes used in this
work. The linear operator, and therefore overall MCSA performance will
be dictated by streaming and scattering as defined on the left-hand
side. We refer the reader to \cite{hamilton_2014} for an extensive
presentation of the $SP_N$ discretization used in this work.

\subsection{Fuel Assembly Model Problem}
\label{subsec:fuel_assembly}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
bottom is given in Figure~\ref{fig:problem3_end}. On the top and
bottom of the assembly, vacuum conditions are used as well as on the
top and right boundaries in
Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=6.5in]{problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

\clearpage

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
In this section we briefly review the Monte Carlo method for linear
systems and Monte Carlo Synthetic Acceleration. Given a linear
problem:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
\mathbb{R}^N$. To form the Monte Carlo method we define the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
following adjoint inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
We can acquire the adjoint solution by forming the Neumann series by
writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence by forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

In addition to the collision estimator, an additional estimator is
available due to the work of Okten \cite{okten_solving_2005} that uses
the method of expected values as a means to improve the Monte Carlo
estimate. As outlined by Spanier and Gelbard
\cite{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator} for a random walk of $k$ events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potential states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to the
collision estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore an unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

At each eigenvalue iteration, we then solve the linear system with the
MCSA:
\begin{subequations}
  \begin{gather}
    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
    \mathbf{\Phi}_{l+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the $l$ subscript indicates the MCSA iteration number and $n$
again indicates the eigenvalue iteration. We refer the reader to
\cite{evans_monte_2014} for more a more complete definition of the
Monte Carlo method for linear systems and MCSA.

@@

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2014} and therefore we apply it here to the $SP_N$
equations given their similar diffusion-like form. A single energy
group problem was first computed with $SP_1$ discretization,
effectively giving the one-speed neutron diffusion system for the fuel
assembly resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
